# Assignment 5: Data analysis

```{r}
library(readr)
human <- read_csv("https://raw.githubusercontent.com/KimmoVehkalahti/Helsinki-Open-Data-Science/master/datasets/human2.csv")
```

## 1 Data

Move the country names to rownames (see Exercise 5.5). Show a graphical overview of the data and show summaries of the variables in the data. Describe and interpret the outputs, commenting on the distributions of the variables and the relationships between them. **(0-3 points)**

```{r}

library(GGally)

# Move the country names to rownames
library(tibble)
human_ <- column_to_rownames(human, "Country")
head(human_)

# graphical overview 
ggpairs(human_, progress = FALSE)

#show summaries of the variables in the data.
summary(human_)
str(human_)
glimpse(human_)

# Describe and interpret the outputs, commenting on the distributions of the variables and the relationships between them. 

# Access corrplot
library(corrplot)

# compute the correlation matrix and visualize it with corrplot
cor(human_)
cor(human_) |> corrplot()

```

## 2. PCA

Perform principal component analysis (PCA) on the raw (non-standardized) human data. Show the variability captured by the principal components. Draw a biplot displaying the observations by the first two principal components (PC1 coordinate in x-axis, PC2 coordinate in y-axis), along with arrows representing the original variables. **(0-2 points)**

```{r}

library(GGally)

pca_human_non_standdd <- prcomp(human_)
pca_human_non_standdd

# draw a biplot of the principal component representation and the original variables
biplot(pca_human_non_standdd, choices = 1:2)

biplot(pca_human_non_standdd, choices = 1:2, cex = c(0.8, 1))
biplot(pca_human_non_standdd, choices = 1:2, cex = c(0.40, 0.60), col = c("grey40", "deeppink2")) # latter affects on vectors
biplot(pca_human_non_standdd, choices = 1:2, cex = c(0.20, 0.60), col = c("grey40", "deeppink2")) # latter affects on vectors
```

## 3. PCA w Standardized variables

Standardize the variables in the human data and repeat the above analysis.

```{r}

human_stddd <- scale(column_to_rownames(human, "Country"))
pca_human_stddd <- prcomp(human_stddd)

# draw a biplot of the principal component representation and the original variables
biplot(pca_human_stddd, choices = 1:2)

biplot(pca_human_stddd, choices = 1:2, cex = c(0.40, 0.60), col = c("grey40", "deeppink2")) # latter affects on vectors

### compare pca_human_non_standdd and pca_human_stddd
biplot(pca_human_non_standdd, choices = 1:2, cex = c(0.20, 0.60), col = c("grey40", "deeppink2")) # latter affects on vectors
biplot(pca_human_stddd, choices = 1:2, cex = c(0.20, 0.60), col = c("grey40", "deeppink2")) # latter affects on vectors
```

Interpret the results of both analysis (with and without standardizing). Are the results different? Why or why not? Include captions (brief descriptions) in your plots where you describe the results by using not just your variable names, but the actual phenomena they relate to. **(0-4 points)**

Interpretation:

## 4. Personal interpretations of the first two principal component dimensions

Give your personal interpretations of the first two principal component dimensions based on the biplot drawn after PCA on the standardized human data. **(0-2 points)**

## 5. Tea & FactoMineR

The tea data comes from the FactoMineR package and it is measured with a questionnaire on tea: 300 individuals were asked how they drink tea (18 questions) and what are their product's perception (12 questions). In addition, some personal details were asked (4 questions).\
Load the tea dataset and convert its character variables to factors:

    tea <- read.csv("https://raw.githubusercontent.com/KimmoVehkalahti/Helsinki-Open-Data-Science/master/datasets/tea.csv", stringsAsFactors = TRUE)

\
Explore the data briefly: look at the structure and the dimensions of the data. Use **View(tea)** to browse its contents, and visualize the data.

```{r}

library(dplyr)
library(tidyr)
library(ggplot2)


# Explore the data briefly: look at the structure and the dimensions of the data. Use View(tea) to browse its contents
tea <- read.csv("https://raw.githubusercontent.com/KimmoVehkalahti/Helsinki-Open-Data-Science/master/datasets/tea.csv", stringsAsFactors = TRUE)
summary(tea)
str(tea)
dim(tea)
glimpse(tea)
View(tea)

# visualize the data.

#pivot_longer(tea, cols = everything(-)) %>% ggplot(aes(value)) + facet_wrap("name", scales = "free") +  geom_bar()

# pivot_longer(tea, cols = everything()) %>%  ggplot(aes(value)) + facet_wrap("name", scales = "free") +   geom_bar() + theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 8))

# error message from previous two functions: filter age variable out

teatea <- dplyr::select(tea, -age)
str(teatea)

pivot_longer(teatea, cols = everything()) %>% 
  ggplot(aes(value)) + facet_wrap("name", scales = "free") +  geom_bar()

pivot_longer(teatea, cols = everything()) %>%  ggplot(aes(value)) + facet_wrap("name", scales = "free") +   geom_bar() + theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 8))




```

Have to say that barplottin every variable (except age) gave kind a messy image. I had to pop up the image and use scale out (cmd + - -command with Mac ) quite a lot. Probably it is easiest to do this assignment using same variables used in Exercise because I cannot find any description about whole data set and variables (although many of factors are quite self-explanatory ex. variable sophisticated with levels "Not.sophisticated" and "sophisticated")

The knitted HTML-version last geom_bar plot was terrible so I'll try again with fewer variables.

```{r}

library(ggplot2)

# column names to keep in the dataset & creation of new a dataset
keep_columns <- c("Tea", "How", "how", "sugar", "where", "lunch")
tea_time <- dplyr::select(tea, keep_columns)

pivot_longer(tea_time, cols = everything()) %>%  ggplot(aes(value)) + facet_wrap("name", scales = "free") +   geom_bar() + theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 8))
```

**MCA.** Use Multiple Correspondence Analysis (MCA) on the tea data (or on just certain columns of the data, it is up to you!). Interpret the results of the MCA. You can also explore other plotting options for MCA. Comment on the output of the plots.

```{r}

library(FactoMineR)
library(swirl)
library(dplyr)

# column names to keep in the dataset & creation of new a dataset
keep_columns <- c("Tea", "How", "how", "sugar", "where", "lunch")
tea_time <- dplyr::select(tea, keep_columns)

# multiple correspondence analysis
mca <- MCA(tea_time, graph = FALSE)

# summary of the model
summary(mca)

# visualize MCA
plot(mca, invisible=c("ind"), graph.type = "classic", habillage = "quali")

## for curiousity using MCA for the whole data (only age variable excluded)

mca2 <- MCA(teatea, graph = FALSE)

# summary of the model (whole data)
summary(mca2)

# visualize MCA (whole data)
plot(mca2, invisible=c("ind"), graph.type = "classic", habillage = "quali")

# in interpration I focus on the first MCA factor map. 
```

Interpretation. Both dimensions explain variance not so good because first dim explains 15% of variance and second 14% variance.

As factor variables, the tea shop as a place and unpacked as a tea product is contribute strongly in dim1.

So based on this analysis, it would be good follow-up question to look if green unpacked tea from tea shop is clear dimension or consumer choice in this data.

Another dimension is contributed by 1) other (with what tea is consumed), 2) chain store + tea shop (where), 3) tea bag + unpacked (how is consumed).

I wonder if these could represent spesific taste of consumers. Dim1 would characterize this hardcore unpacked green tea consumer and Dim2 this consumer type who is more open to different ways to consume tea.

look support from here:

<http://factominer.free.fr/factomethods/multiple-correspondence-analysis.html>

Youtube: <https://www.youtube.com/watch?v=reG8Y9ZgcaQ>

... with

"Draw at least the variable biplot of the analysis." No idea what should be done.
