# Assignment 4. **Analysis exercises** (Max 15 points)

## 1.The data

Explore the structure and the dimensions of the Boston data and describe the dataset briefly, assuming the reader has no previous knowledge of it. Details about the Boston dataset can be seen for example [here](https://stat.ethz.ch/R-manual/R-devel/library/MASS/html/Boston.html). **(0-1 points)**

The Housing Values in Suburbs of Boston.

-   Dataset contains 14 colums including crime rate (mean 3.6, median 0.3), pupil-teacher ratio (mean 18.46, median 19.05) and non-retail business acres (mean 408, median 330) per town. Other interesting columns are for example distance from Boston center (mean 3.795, median 3.207), property tax-rate (mean 408.2, median 330) and amount of population of lower status (mean 12.65, median 11.36). There is no missingness in the data and rows 506 (towns?) in the data. All the data contain numerical variables, chas is binary.

-   Full list of columns are following (from [here](https://stat.ethz.ch/R-manual/R-devel/library/MASS/html/Boston.html)).

    -   crim per capita crime rate by town.

    -   zn proportion of residential land zoned for lots over 25,000 sq.ft.

    -   indus proportion of non-retail business acres per town.

    -   chas Charles River dummy variable (= 1 if tract bounds river; 0 otherwise).

    -   nox nitrogen oxides concentration (parts per 10 million).

    -   rm average number of rooms per dwelling.

    -   age proportion of owner-occupied units built prior to 1940.

    -   dis weighted mean of distances to five Boston employment centres.

    -   rad index of accessibility to radial highways.

    -   tax full-value property-tax rate per \$10,000.

    -   ptratio pupil-teacher ratio by town.

    -   black 1000(Bk−0.63)21000(Bk−0.63)2 where BkBk is the proportion of blacks by town.

    -   lstat lower status of the population (percent).

-   medv median value of owner-occupied homes in \$1000s.

```{r}

#install.packages("MASS")
library(MASS)
library(finalfit)
library(dplyr)
library(corrplot)
data("Boston")
glimpse(Boston)
ff_glimpse(Boston)
summary(Boston)

```


## 2. Graphical overview

Show a graphical overview of the data and show summaries of the variables in the data. Describe and interpret the outputs, commenting on the distributions of the variables and the relationships between them. **(0-2 points)**

**Ggpairs.** First I tried to the ggpairs plot matrix from previous weeks. The image is so bad that I had hard time to figuring out whats going on there. I added "proportions = auto" which helped little. Ggpairs show interesting correlations. For example variables listed there is positive correlation between crime rate and low status population ratio and negative correlation between crime rate and population of afro-americans.

**Pairs.** This plot gives same information than the last one. Ggpairs is bit more helpful when there is the correlation indicator.

**Corrplot.** This is gives same information than Ggpairs. Probably I will use corrplot in future than ggpairs because this is more tidy and it is easier to see which variables have higher correlation. For example dis (distanse from center) has negative correlation between indus (proportion of non-retail business acres), nox (itrogen oxides concentration) and age ( proportion of owner-occupied units built prior to 1940). Crime rate has positive correlations with rad (index of accessibility to radial highways) and tax (full-value property-tax rate per \$10,000).

```{r}
library(GGally) 
library(ggplot2)

# a plot matrix with ggpairs()

p2 <- ggpairs(Boston, mapping = aes(alpha = 0.3), lower = list(combo = wrap("facethist", bins = 20)), proportions = "auto")
p2

pairs(Boston)

# calculate the correlation matrix and round it
cor_matrix <- cor(Boston) 

# print the correlation matrix
cor_matrix

# visualize the correlation matrix
library(corrplot)
corrplot(cor_matrix, method="circle")


```

## 3. Standardize the dataset

Standardize the dataset and print out summaries of the scaled data. How did the variables change? Create a categorical variable of the crime rate in the Boston dataset (from the scaled crime rate). Use the quantiles as the break points in the categorical variable. Drop the old crime rate variable from the dataset. Divide the dataset to train and test sets, so that 80% of the data belongs to the train set. **(0-2 points)**

**Standardize & scale the dataset. How did the variables change?** First I notice that crime rate dropped from mean 3.61 and median 0.25651 to mean 0 and median -0.390280. Max values in every variable dropped significantly.

I tried corrplot out of curiosity and see no changes (no \*\*\*\*, sherlock).

```{r}
# center and standardize variables
boston_scaled <- as.data.frame(scale(Boston))
boston_scaled$crim <- as.numeric(boston_scaled$crim)
  
# summaries of the scaled variables
summary(Boston)
summary(boston_scaled)
glimpse(boston_scaled)

#cor_matrix2 <- cor(boston_scaled) 
#cor_matrix2
#corrplot(cor_matrix2, method="circle")
```

Create a categorical variable of the crime rate in the Boston dataset (from the scaled crime rate). Use the quantiles as the break points in the categorical variable.

```{r}

bins <- quantile(boston_scaled$crim)
bins

# created a categorical variable 'crime': low, med_low, med_high and high
crime <- cut(boston_scaled$crim, breaks = bins, labels = c("low", "med_low", "med_high", "high"), include.lowest = TRUE)

# looking at the table of the new factor crime
crime

```

Drop the old crime rate variable from the dataset

```{r}
# remove original crim from the dataset
boston_scaled <- dplyr::select(boston_scaled, -crim)

# add the new categorical value to scaled data
boston_scaled <- data.frame(boston_scaled, crime)

summary(boston_scaled)
```

Divide the dataset to train and test sets, so that 80% of the data belongs to the train set.

```{r}
# number of rows in the Boston dataset 
n <- nrow(boston_scaled)

# choose randomly 80% of the rows
ind <- sample(n,  size = n * 0.8)

# create train set
train <- boston_scaled[ind,]

# create test set 
test <- boston_scaled[-ind,]
```

## 4. Linear discriminant analysis

Fit the linear discriminant analysis on the train set. Use the categorical crime rate as the target variable and all the other variables in the dataset as predictor variables. Draw the LDA (bi)plot. **(0-3 points)**

```{r}
# linear discriminant analysis
lda.fit <- lda(crime ~ ., data = train)

# print the lda.fit object
lda.fit

# the function for lda biplot arrows
lda.arrows <- function(x, myscale = 1, arrow_heads = 0.1, color = "red", tex = 0.75, choices = c(1,2)){
  heads <- coef(x)
  graphics::arrows(x0 = 0, y0 = 0, 
         x1 = myscale * heads[,choices[1]], 
         y1 = myscale * heads[,choices[2]], col=color, length = arrow_heads)
  text(myscale * heads[,choices], labels = row.names(heads), 
       cex = tex, col=color, pos=3)
}

# target classes as numeric
classes <- as.numeric(train$crime)

# plot the lda results (select both lines and execute them at the same time!)
plot(lda.fit, dimen = 2)
lda.arrows(lda.fit, myscale = 1)


```

## 5. Predict & cross tabulate

Save the crime categories from the test set and then remove the categorical crime variable from the test dataset. Then predict the classes with the LDA model on the test data. Cross tabulate the results with the crime categories from the test set. Comment on the results. **(0-3 points)**

```{r}
correct_classes <- test$crime
test <- dplyr::select(test, -crime)

# predict classes with test data
lda.pred <- predict(lda.fit, newdata = test)

# cross tabulate the results
table(correct = correct_classes, predicted = lda.pred$class)

```

## 6. Standardize the dataset

Reload the Boston dataset and standardize the dataset (we did not do this in the Exercise Set, but you should scale the variables to get comparable distances).

```{r}

data("Boston")

# center and standardize variables
boston_scaled <- scale(Boston)
  
# summaries of the scaled variables
summary(Boston)
summary(boston_scaled)

# class of the boston_scaled object
class(boston_scaled)

# change the object to data frame
boston_scaled <- as.data.frame(boston_scaled)
```

Calculate the distances between the observations.

```{r}
# euclidean distance matrix
dist_eu <- dist(Boston)

# look at the summary of the distances
summary(dist_eu)

# manhattan distance matrix
dist_man <- dist(Boston, method = "manhattan")

# look at the summary of the distances
summary(dist_man)

```

Run k-means algorithm on the dataset.

```{r}
km <- kmeans(Boston, centers = 4)

# plot the Boston dataset with clusters
pairs(Boston, col = km$cluster)

# k-means clustering
km <- kmeans(Boston, centers = 4)

# plot the Boston dataset with clusters
pairs(Boston[6:10], col = km$cluster)

####

# k-means clustering
km <- kmeans(Boston, centers = 3)

# plot the Boston dataset with clusters
pairs(Boston[c("rm", "age", "dis", "crim")], col = km$cluster)
```

Investigate what is the optimal number of clusters and run the algorithm again.

```{r}
set.seed(123)

# determine the number of clusters
k_max <- 10

# calculate the total within sum of squares
twcss <- sapply(1:k_max, function(k){kmeans(Boston, k)$tot.withinss})

# visualize the results
qplot(x = 1:k_max, y = twcss, geom = 'line')

# k-means clustering
km <- kmeans(Boston, centers = 10)

# plot the Boston dataset with clusters
pairs(Boston, col = km$cluster)

```

The optimal number of clusters is when the total WCSS drops radically. In this example twcss drops when amount of clusters is two. I run the algorithm again with this.

```{r}
k_max <- 2

# calculate the total within sum of squares
twcss <- sapply(1:k_max, function(k){kmeans(Boston, k)$tot.withinss})

```

Visualize the clusters (for example with the pairs() or ggpairs() functions, where the clusters are separated with colors) and interpret the results. **(0-4 points)**

```{r}
# visualize the results
qplot(x = 1:k_max, y = twcss, geom = 'line')

# k-means clustering
km <- kmeans(Boston, centers = 2)

# plot the Boston dataset with clusters
pairs(Boston, col = km$cluster)
```
