### Assignment 5: Data analysis

```{r}
library(readr)
human <- read_csv("https://raw.githubusercontent.com/KimmoVehkalahti/Helsinki-Open-Data-Science/master/datasets/human2.csv")
```

## 1 Data

Move the country names to rownames (see Exercise 5.5). Show a graphical overview of the data and show summaries of the variables in the data. Describe and interpret the outputs, commenting on the distributions of the variables and the relationships between them. **(0-3 points)**

```{r}

library(GGally)

# Move the country names to rownames
library(tibble)
human_ <- column_to_rownames(human, "Country")
head(human_)

# graphical overview 
ggpairs(human_, progress = FALSE)

#show summaries of the variables in the data.
summary(human_)
str(human_)
glimpse(human_)

# Describe and interpret the outputs, commenting on the distributions of the variables and the relationships between them. 

# Access corrplot
library(corrplot)

# compute the correlation matrix and visualize it with corrplot
cor(human_)
cor(human_) |> corrplot()

```

## 2. PCA

Perform principal component analysis (PCA) on the raw (non-standardized) human data. Show the variability captured by the principal components. Draw a biplot displaying the observations by the first two principal components (PC1 coordinate in x-axis, PC2 coordinate in y-axis), along with arrows representing the original variables. **(0-2 points)**

```{r}

library(GGally)

pca_human_non_standdd <- prcomp(human_)
pca_human_non_standdd

# draw a biplot of the principal component representation and the original variables
biplot(pca_human_non_standdd, choices = 1:2)

biplot(pca_human_non_standdd, choices = 1:2, cex = c(0.8, 1))
biplot(pca_human_non_standdd, choices = 1:2, cex = c(0.40, 0.60), col = c("grey40", "deeppink2")) # latter affects on vectors
biplot(pca_human_non_standdd, choices = 1:2, cex = c(0.20, 0.60), col = c("grey40", "deeppink2")) # latter affects on vectors
```

## 3. PCA w Standardized variables

Standardize the variables in the human data and repeat the above analysis.

```{r}

human_stddd <- scale(column_to_rownames(human, "Country"))
pca_human_stddd <- prcomp(human_stddd)
pca_human_stddd

# draw a biplot of the principal component representation and the original variables
biplot(pca_human_stddd, choices = 1:2)

biplot(pca_human_stddd, choices = 1:2, cex = c(0.40, 0.60), col = c("grey40", "deeppink2")) # latter affects on vectors

### compare pca_human_non_standdd and pca_human_stddd
biplot(pca_human_non_standdd, choices = 1:2, cex = c(0.20, 0.60), col = c("grey40", "deeppink2")) # latter affects on vectors
biplot(pca_human_stddd, choices = 1:2, cex = c(0.20, 0.60), col = c("grey40", "deeppink2")) # latter affects on vectors
```

Interpret the results of both analysis (with and without standardizing). Are the results different? Why or why not? Include captions (brief descriptions) in your plots where you describe the results by using not just your variable names, but the actual phenomena they relate to. **(0-4 points)**

Interpretation:

## 4. Personal interpretations of the first two principal component dimensions

Give your personal interpretations of the first two principal component dimensions based on the biplot drawn after PCA on the standardized human data. **(0-2 points)**

## 5. Tea & FactoMineR

The tea data comes from the FactoMineR package and it is measured with a questionnaire on tea: 300 individuals were asked how they drink tea (18 questions) and what are their product's perception (12 questions). In addition, some personal details were asked (4 questions).\
Load the tea dataset and convert its character variables to factors:

    tea <- read.csv("https://raw.githubusercontent.com/KimmoVehkalahti/Helsinki-Open-Data-Science/master/datasets/tea.csv", stringsAsFactors = TRUE)

\
Explore the data briefly: look at the structure and the dimensions of the data. Use **View(tea)** to browse its contents, and visualize the data.\
Use Multiple Correspondence Analysis (MCA) on the tea data (or on just certain columns of the data, it is up to you!). Interpret the results of the MCA and draw at least the variable biplot of the analysis. You can also explore other plotting options for MCA. Comment on the output of the plots. \*\*(0-4 points)\*\*

**After completing all the phases above you are ready to submit your Assignment for the review (using the Moodle Workshop below).**

**Have the two links (your GitHub repository and your course diary) ready!**
