# Week 6: Analysis of longitudinal data   

As always I ran short of time. Unfortunately this time I did not make it to part the end of Part I and could not start in Part II.

## Meet and Repeat: PART I

-   Print out the (column) names of the data
-   Look at the structure of the data
-   Print out summaries of the variables in the data
-   Pay special attention to the structure of the data

```{r}

RATS_ <- read.table("https://raw.githubusercontent.com/KimmoVehkalahti/MABS/master/Examples/data/rats.txt", sep ="\t", header = T)

names(RATS_)

str(RATS_)

summary(RATS_)
```

###  pivot_longer()

```{r}
library(dplyr)
library(tidyr)

# Factor treatment & subject

RATS_$ID <- factor(RATS_$ID)
RATS_$Group <- factor(RATS_$Group)

# Convert to long form

RATSL_ <- pivot_longer(RATS_, cols = -c(ID, Group), 
                      names_to = "WD",
                      values_to = "Weight") %>% 
  mutate(Time = as.integer(substr(WD, 3, 4))) %>%
  arrange(Time)

# Take a glimpse at the BPRSL data
glimpse(RATSL_)

```

### Individuals on the plot

```{r}
#Access the package ggplot2
library(ggplot2)

# Draw the plot
ggplot(RATSL_, aes(x = Time, y = Weight, linetype = ID)) +
  geom_line() +
  scale_linetype_manual(values = rep(1:10, times=4)) +
  facet_grid(. ~ Group, labeller = label_both) +
  theme(legend.position = "none") + 
  scale_y_continuous(limits = c(min(RATSL_$Weight), max(RATSL_$Weight)))

```

### Standardise

```{r}
library(dplyr)
library(tidyr)

# Standardise the variable Weight
RATSL_ <- RATSL_ %>%
  group_by(Time) %>%
  mutate(stdWeight = scale(Weight)) %>%
           ungroup(stdWeight)
  
RATSL_$stdWeight <- as.numeric(RATSL_$stdWeight)

# Glimpse the data
glimpse(RATSL_)

# Plot again with the standardised bprs
library(ggplot2)

ggplot(RATSL_, aes(x = Time, y = Weight, linetype = ID)) +
  geom_line() +
  scale_linetype_manual(values = rep(1:10, times=4)) +
  facet_grid(. ~ Group, labeller = label_both) +
  theme(legend.position = "none") + 
  scale_y_continuous(name = "normal bprs")

ggplot(RATSL_, aes(x = Time, y = stdWeight, linetype = ID)) +
  geom_line() +
  scale_linetype_manual(values = rep(1:10, times=4)) +
  facet_grid(. ~ Group, labeller = label_both) +
  theme(legend.position = "none") + 
  scale_y_continuous(name = "standardized bprs")
```

## Meet and Repeat: PART II    

Ran out of time :(
